Highly Available Transactions: Virtues and Limitations

### 2013-03-26 writeup/intro.tex
Another pass at writing

Many recent systems have eschewed ACID guarantees.

CAP Theorem says you can't have consistency and availability and also talks about partitions.

What does CAP have to say about ACID? Recent workshop result on Highly Available Transactions briefly demonstrated that some ACID guarantees are indeed achievable in a highly available manner.

Many ACID systems were designed either i.) for single systems or ii.)
to provide serializability. The combination of weak isolation and
distributed systems necessitates further study and
design. \textit{What is the benefit of high availability and what is
  the cost?}

Highly Available Transactions are the class of multi-object operations
and their respective semantics that can be provided with high
availability.

Weak isolation has a cost: in the words of one database pioneer, you
``fall off a cliff'' when it comes to reasoning about your
program. However, it also has a benefit: availability and latency.

Our goal in this paper is to objectively map the space of trade-offs and

 This paper fills in the remaining gaps.

What does high availability even mean in a transaction environment?

What models are achievable, and which aren't?

@@

\begin{itemize}
\item We present a model for high availability in a transactional environment, including Replica-agnostic High Availability (R-HA) and Sticky High Availability (S-HA).
\item We analyze the existing literature on ACID properties (particularly, ANSI SQL Spec, Adya, Gray) to show which properties* with a focus on isolation* are achievable in HATS. Achievable properties include transactional atomicity, variants of repeatable read isolation, and session guarantees. We also describe unavailable properties, like preventing Lost Update and Read and Write Skew. We believe that this is the first such characterization in the literature.
\item We further motivate the study of HATS via the analysis of existing concurrency control algorithms in a highly available environment as well as through deployment of an experimental database prototype on public cloud infrastructure.
\end{itemize}

>>>>

Requirements for high availability, predictable low latency, and
simplicity of coordination protocols have led many recent distributed
database designers to renounce several guarantees provided by
traditional database systems. In particular, many modern systems forgo
providing ACID guarantees---which often provide single-system image
(SSI) programmability but suffer from unavailability and often
expensive coordination costs. The CAP Theorem, demonstrating the
incompatibility of ``always on'' operation with guarantees on data
item recency in the presence of unreliable networks, has led a large
number of systems to prioritize high availability and guaranteed low
latency, which are especially important in geo-replicated contexts. We
observe a dearth of systems providing general-purpose multi-operation
\textit{transactions} over multiple data items with useful semantics.

The relationship of the CAP Theorem and ACID semantics is poorly
understood. While serializability---the gold standard of SSI
transactions---is unavailable, databases have a long tradition of
providing weaker isolation and consistency guarantees. While these
weaker guarantees do not provide SSI for arbitrary general-purpose
read/write transactions, they often admit greater concurrency and
higher performance than comparable SSI algorithms. Accordingly, weak
isolation is widely employed (Section~\ref{needed}), overwhelmingly as
the default setting in today's traditional ``ACID'' and up-and-coming
``NewSQL'' databases and often as the maximum guarantees offered
(Section~\ref{sec:modernacid}). While the prevalence of these
configurations is encouraging, the current literature does not provide
guidance as to the operation of weakly consistent isolation guarantees
in a distributed context. Rather, we have a strong understanding of
weak isolation in the single-server context from which it originated
as well as several techniques for providing distributed
serializability but we have little to no insight into which ACID
guarantees are compatible with Highly Available Transactions (HATs),
or can be provided with high availability and low latency.

This work attempts to bridge the gap between ACID properties and
highly available systems. We analyze a range of existing ACID and
distributed consistency models to provide a rigorous taxonomization of
which models are achievable in a HAT system.  While several
guarantees, like Serializability and Snapshot Isolation, are
unachievable due to reliance on preventing Lost Update and Write Skew
phenomena, many other ACID guarantees are HAT compliant, such as
providing Read Committed isolation, preventing Fuzzy Read phenomena,
and ensuring atomicity of transactional updates. We also attempt to
bridge the often distant worlds of distributed system consistency and
ACID database consistency by analyzing the relationship between
so-called session guarantees, ``register'' semantics, and ACID
properties. In doing so, we introduce a hierarchy of highly available
properties as, for example, some guarantees require ``stickiness,'' or
fate-sharing between clients and database servers. We believe that
this analysis is the first that has been performed over such a broad
range of ACID and distributed consistency models.

While this model classification is useful, we also seek to understand
the implications for end-users of HAT systems. HATs provide high
availability and low latency but cannot offer all of the semantic
guarantees of unavailable systems---when is this trade-off worthwhile?
We attempt to objectively study these virtues and limitations of both
classic and HAT systems by performing an extensive survey of
practitioner accounts and research literature, performing experimental
analyses on modern cloud infrastructure, and analyzing representative
applications for their semantic requirements. While our results are
not definitive for all applications, they suggest that HATs offer a
one to three order of magnitude latency decrease compared to
traditional protocols and provide acceptable semantics for a wide
range of programs, particularly those with monotonic logic and
commutative updates. HAT systems can also provide arbitrary foreign
key constraints on multi-item updates and sometimes provide uniqeness
guarantees. However, particularly for programs with non-monotonic
logic, HATs may fall short, requiring more coordination-intensive
mechanisms. Concisely, we attempt to understand \textit{what benefits
  do Highly Available Transactions offer and what is their cost?}

@@

\begin{itemize}
\item We present a model for high availability in a transactional
  environment, replica-based High Availability, Sticky High
  Availability, and Transactional High Availability.

\item We analyze the existing literature on ACID properties and
  replicated data semantics to show which properties are achievable in
  HAT systems. Achievable properties include transactional atomicity,
  variants of repeatable read isolation, and session guarantees. We
  also describe unavailable properties, like preventing Lost Update
  and Write Skew. We believe that this is the first such
  characterization in the literature.

\item We further motivate the study of HATs via the analysis of
  existing concurrency control algorithms, a case-study of running a
  transactional application with HATs, and via deployment of a HAT
  database prototype on public cloud infrastructure.
\end{itemize}

### 2013-03-26 writeup/model.tex
Another pass at writing

To understand which guarantees can be provided with high availabilty,
we must first understand what properties define high availability. In
this section, we will formulate a model that captures a range of
availability models, including ``true'' high availability,
availablility with stickiness, and several other useful variants.

Informally, highly available algorithms ensure ``always on'' operation
and promise a guaranteed response. If users of a highly available
system are able to contact a (set of) server(s) in a system, they are
guaranteed a response as the (set of) server(s) does not need to
communicate with any others. This lack of fast-path coordination also
means that a highly available system can also offer low latency
operation~\cite{abadi-pacelc}. To describe whether a transactional
system is highly available, we need both a way to describe what
servers a client must contact as well as what kinds of responses a
server can provide, especially given the possibility of aborts.

@@

Traditionally, a system provides high availability if every user that
can contact a server eventually receives a response from that server,
even in the presence of arbitrary, indefinitely long network
partitions between servers~\cite{gilbert-cap}. This is a useful
definition: if we can contact at least one server, then we can receive
a response. However, there are several other cases we should consider.

First, clients may wish to contact multiple servers. For example, to
ensure that the effects of its operations will persist in the face of
server failures, clients may wish to wait until operations reach more
than one server before they considers operations completed. However,
according to our definition of high availability, server-fault
tolerance is unavailable. To allow flexibility in the number of
servers required for an operation, we propose
\textit{$K$-availability}: if a client can contact $K$ servers, then
it eventually receives a response from all of them, even in the
presence of indefinitely long network partitions. Compared to a system
providing high availability ($1$-availability), a system with
$2$-availability and will require more messaging, possibly higher
latency and more unsuccessful operations in the presence of network
and server failures.  Additionally, as is commmon in standard
consensus algorithms like Paxos, clients may want to contact a
majority of $N$ servers in a system. This \textit{majority
  availability} (a special case of $K$-availability with $K=\lceil
\frac{N}{2} \rceil$) is more available than $N$-availability.

Clients may also wish to contact the same server (or set of servers)
across subsequent operations. As we will discuss in
Section~\ref{sec:hats}, clients can ensure continuity between
operations (e.g., reading their prior updates to a data item) by
maintaining affinity or ``stickiness'' with a server or set of
servers~\cite{vogels-defs}. Simultaneously aintaining availability and
stickiness requires fate-sharing between the client and its set of
servers: if a client's sticky servers fail, or, equivalently, if the
client is partitioned from its sticky servers, then it may face
unavailability. We say that a system provides \textit{sticky
  availability} if a client's subsequent operations are executed
against a server that has observed all of its prior operations, then
it eventually receives a response, even in the presence of
indefinitely long partitions. As above, it may be useful for a client
to achieve stickiness \textit{and} contact many replicas (i.e., for
fault tolerance), so we generalize sticky availability to
\textit{$K$-sticky availability}, which provides sticky availability
with $K$ distinct servers. A client may choose to become $1$-sticky
available by acting as a server itself; for example, a client might
cache its reads and writes. However, a single client cannot achieve
$K$-sticky availability with $K>1$ on its own. Moreover, ``sticking''
all clients to a designated set of servers (i.e., master availability)
is a special case of sticky availability and is less available than
the equivalent sticky availability.

@@

\end{tikzpicture}
\label{fig:availability-order}
\caption{Hierarchy of replica availability levels for $N>3$ servers.}
\end{figure}

We show the hierarchy of replica availabilty levels in
Figure~\ref{fig:availability-order}. $K$-sticky availability subsumes
$K$-availability, but $K$-sticky availability is incomparable with
$(K+1)$-availability. $N$-sticky availability is equivalent to
$N$-availability, while majority availability subsumes $1$-sticky
availability More generally, $\lceil \frac{N}{2} \rceil$$+$$K$$-$$1$
availability subsumes $K$-sticky availability. We have omitted
discussion of operation-specific availability levels (e.g.,
$N$-availability for writes and $1$-availability for reads in a
write-all, read-one data store), system membership changes, or
heterogeneous replicas (e.g., servers in a local and remote
datacenters) but believe there are several avenues for further
taxnomization.

@@

Thus far, we have focused on single-operation availability. This is
standard in distributed systems literature (e.g., atomic, regular, and
safe register models all concern single objects~\cite{herlihy-art}),
yet the database literature largely focuses on transactions: groups of
multiple operations over multiple objects. Accordingly, by itself,
replica availability is insufficient to describe availability
guarantees for transactions. Additionally, given the choice of
\textit{commit} and \textit{abort} responses---which signal
transaction success or failure to a client---we must be careful in
defining availability.

If a client wishes to execute a transaction that performs operations
on multiple data items, then the client must be able to contact and
receive a response from at least one server responsible for each data
item. This may result in lower availability than a non-transactional
requirement. Additionally, given the possiblity of aborts, we need to
ensure useful forward progress. A system can trivially guarantee
clients a response by always aborting their
transactions~\cite{transaction-liveness}. However, this is an
unsatisfactory system because nothing good (transaction commit) ever
happens; instead, we require a \textit{liveness} property. We cannot
guarantee that every transaction will commit---transactions may choose
to abort themselves---but we need to make sure that the system will
not indefinitely abort transactions on its own volition. We call a
transaction abort due to a transaction's own choosing (e.g., as part
of the transaction itself or due to a would-be integrity constraint
violation) an \textit{internal abort} and an abort due to system
implementation an \textit{external abort}.

We say that a system provides \textit{transactional availability} if,
given replica availability for every data item in a transaction, the
transaction eventually commits or internally aborts. For example,
system will violate transactional $2$-availability if client that can
contact two servers for each of data items in its transaction does not
eventually commit in the absence of internal aborts.

>>>>

To understand which guarantees can be provided with high availabilty,
we must first define what high availability means. In this section, we
will formulate a model that captures a range of availability models,
including ``true'' high availability, availablility with stickiness,
and several other useful variants.

Informally, highly available algorithms ensure ``always on'' operation
and guaranteed low latency. If users of a highly available system are
able to contact a (set of) server(s) in a system, they are guaranteed
a response as the (set of) server(s) does not need to communicate with
any others. If servers are partitioned from one another, they do not
need to stall in order to provide a ``safe'' query response. This lack
of fast-path coordination also means that a highly available system
also provides low latency~\cite{abadi-pacelc}. However, to properly
describe whether a \textit{transactional} system is highly available,
we need both a way to describe what servers a client must contact as
well as what kinds of responses a server can provide, especially given
the possibility of aborts.

@@

Traditionally, a system provides high availability if every user that
can contact a server eventually receives a response from that server,
even in the presence of arbitrary, indefinitely long network
partitions between servers~\cite{gilbert-cap}. As in a standard
distributed database, designated servers might handle operations on
different data items;\footnote{There is a further distinction between a
  fully replicated system, in which all servers are replicas for all
  data items, and a partially replicated system, in which at least one
  server acts as a replica for only a subset of all data items. For
  generality, and, for applicability to many modern ``sharded'' or
  ``partitioned'' data storage systems like Dynamo, we consider
  partially replicated systems.} a server that can handle an
operation for a given data item is called a \textit{replica} for that
item. This is a useful definition: if a client can contact at least
one replica for an item on which it wishes to perform an operation,
then the replica will provide a response to the client. However, there
are several other cases we should consider.

First, clients may wish to contact multiple servers. For example, to
ensure that the effects of operations will persist in the face of
server failures, clients may wish to wait until operations reach more
than one server before they consider operations completed. However,
according to our definition of high availability, server-fault
tolerance is unavailable. To allow flexibility in the number of
servers required for an operation, we propose
\textit{$K$-availability}: if a client can contact $K$ servers, then
it eventually receives a response from all of them, even in the
presence of indefinitely long network partitions. Compared to a system
providing high availability ($1$-availability), a system with
$2$-availability will require more messaging and possibly higher
latency and may experience more unsuccessful operations in the
presence of network and server failures.  Additionally, as is commmon
in standard consensus algorithms like Paxos, clients may want to
contact a majority of $N$ servers in a system. This \textit{majority
  availability} (a special case of $K$-availability with $K=\lceil
\frac{N}{2} \rceil$) is more available than full $N$-availability.

Clients may also wish to contact the same server (or set of servers)
across subsequent operations. As we will discuss in
Section~\ref{sec:hats}, clients can ensure continuity between
operations (e.g., reading their prior updates to a data item) by
maintaining affinity or ``stickiness'' with a server or set of
servers~\cite{vogels-defs}. Simultaneously maintaining availability
and stickiness requires fate-sharing between the client and its set of
servers: if a client's sticky servers fail, or, equivalently, if the
client is partitioned from its sticky servers, then it may face
unavailability. We say that a system provides \textit{sticky
  availability} if a client's subsequent operations are executed
against a server that has observed all effects of its prior
operations, then it eventually receives a response, even in the
presence of indefinitely long partitions. As above, it may be useful
for a client to achieve stickiness \textit{and} contact many replicas
(i.e., for fault tolerance), so we generalize sticky availability to
\textit{$K$-sticky availability}, which provides sticky availability
with $K$ distinct servers. A client may choose to become $1$-sticky
available by acting as a server itself; for example, a client might
cache its reads and writes. However, a single client cannot achieve
$K$-sticky availability with $K>1$ on its own. Moreover, ``sticking''
all clients to a designated set of servers (i.e., master availability)
is a special case of sticky availability and is less available than
the equivalent non-mastered sticky availability.

@@

\end{tikzpicture}
\caption{Hierarchy of replica availability levels for $N>3$ servers.}
\label{fig:availability-order}
\end{figure}

We show a hierarchy of replica availability levels in
Figure~\ref{fig:availability-order}. $K$-sticky availability subsumes
$K$-availability, but $K$-sticky availability is incomparable with
$(K+1)$-availability. $N$-sticky availability is equivalent to
$N$-availability, while majority availability subsumes $1$-sticky
availability. More generally, $\lceil \frac{N}{2} \rceil$$+$$K$$-$$1$
availability subsumes $K$-sticky availability. We have omitted
discussion of operation-specific availability levels (e.g.,
$N$-availability for writes and $1$-availability for reads in a
write-all, read-one data store), system membership changes, or
heterogeneous replicas (e.g., servers in a local and remote
datacenters) but believe there are several avenues for further
taxnomization.

@@

So far, we have focused on single-object, single-operation
availability. This is standard in distributed systems literature
(e.g., atomic, regular, and safe register models all concern single
objects~\cite{herlihy-art}), yet the database literature largely
focuses on transactions: groups of multiple operations over multiple
objects. Accordingly, by itself, replica availability is insufficient
to describe availability guarantees for transactions. Additionally,
given the choice of \textit{commit} and \textit{abort}
responses---which signal transaction success or failure to a
client---we must be careful in how we define availability.

If a client wishes to execute a transaction that performs operations
on multiple data items, then the client must be able to contact and
receive a response from at least one replica for each data item. This
may result in lower availability than a non-transactional
requirement. Additionally, given the possiblity of aborts, we need to
ensure useful forward progress. A system can trivially guarantee
clients a response by always aborting their
transactions~\cite{transaction-liveness}. However, this is an
unsatisfactory system because nothing good (transaction commit) ever
happens; instead, we require a \textit{liveness} property. We cannot
guarantee that every transaction will commit---transactions may choose
to abort themselves---but we need to make sure that the system will
not indefinitely abort transactions on its own volition. We call a
transaction abort due to a transaction's own choosing (e.g., as part
of the transaction itself or due to a would-be integrity constraint
violation) an \textit{internal abort} and an abort due to system
implementation an \textit{external abort}.

We say that a system provides \textit{transactional availability} if,
given replica availability for every data item in a transaction, the
transaction eventually commits or internally aborts~\cite{hat-hotos}. For example,
system will violate transactional $2$-availability if client that can
contact two servers for each of data items in its transaction does not
eventually commit in the absence of internal aborts.
